Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-05-28 manuscript/manuscript.qmd
model 6 corrections

This model follows the OC6 formulation proposed by O’Reilly et al. (2019), with the difference that the denominator is the sum of $Rrs555$ and $Rrs670$ rather than the mean. The model mean is a fourth-order polynomial of $log(MBR)$ to estimate $log(Chl)$. The set of proabilistic equations that define this model can be found in the Supplemental Material Section. The model structure and inference flow is represented in @fig-model1_struct as a Directed Acyclic Graph (DAG)

::: {#fig-oc6-struct pos=H}

![](images/model1_structure.png)

Model 1 Directed Acyclic Graph showing model parameters and their prior distributions and how they relate . The intercept $α$ and the polynomial slopes, $β1$-$β4$ all have Gaussian priors; $σ$, the likelihood dispersion parameter, receives a Gamma prior; the likelihood itself is constrained by a Truncated Normal distribution (see text for more). The plate containing the likelihood is indexed ($obs\_idx$) indicating the likelihood will be computed for each of the 1082 observations contained in the dataframe and show the number of data points available for fitting.

@@

::: {#fig-model2-struct}

![](images/model2_structure.png)

Model 2 - Multi-level hierarchical level with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510). Model parameter prior mean and variance have Gaussian and Exponential hyperpriors. This structure allows the sharing of information across groups at model fitting time.

#### Model 5: HLR with Added Linear Heteroscedastic Likelihood Dispersion

The data plotted in @fig-eda suggests heteroscedasticity - variance that is not constant, but rather changes depending on the data. Models 3 and 4 investigate the role of group-wise constant heteroscedasticity and global linear heteroscedasticity, respectively. Details of these models can be found in the Supplementary Material section. To investigate this further, Model 5 looks at group-wise linear heteroscedasticity. Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood. Similar to the model mean,  $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors. Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.  illustrates the model structure.

::: {#fig-model5-struct}

![](images/model5_structure.png)

Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.  Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2.
:::

####  Model 6:  Augmenting Model 5 with Measurement-Specific Dispersion

A reasonable assumption is that HPLC measurements are less noisy than fluorescence. This model adds a measurement-method-specific dispersion term to the structure of Model 6, to capture the suspected increase in variability associated with fluorescence-based chlorophyll estimates. A reasonable assumption is that HPLC measurements are less noisy than fluorescence. The likelihood dispersion, $σ$, is given the expression $log(σ) = α_j + β_j × log(MBR) + γ_{chl\_type} * (1 - chl_{type\_idx})$, where $chl_{type\_idx} = 1$ if *in-situ* chlorophyll was measured with HPLC, $0$ otherwise. As a result, $σ$ is the same as in Model 5 if $chl_{type\_idx} = 1$, since the extra term is nullified. In the case of fluorescence measurement, on the other hand, $log(σ)$ receives an extra noise term, providing an effective measurement of noise difference between the two methods.

@fig-model6-struct shows the inference DAG of Model 6.

::: {#fig-model6-struct fig-pos="H"}

@@

*Priors* - I used weakly informative priors across all models to balance flexibility with regularization.  For regression parameters  I used Gaussian priors, In the case of the  single-level  Model 1, this meant priors with $0$ mean and variance of $1$ to encourage parameter sparsity. In the case of Models 2-6, which are multi-level models, hyperpriors of prior means were set to Gaussans of mean 0 and variance 1, again to encourage sparsity. Hyperpriors of prior variances were  Exponential distributions with rate set to $1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise. This hyperprior also encourages uncertainty shrinkage across groups if possible. In the case of homoscedasticity, or group-wise heteroscedasticity, I used a Gamma prior of shape and scale of $2$ for the likelihood dispersion $σ$, which provides strictly postivie support. Prior soundness was assessed using *Prior Predictive Checks*. This procedure takes advantage of the fact taht Bayesian models are generative, meaning the model can be sampled without data, to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting. Prior predictive $log(Chl)$ values were plotted as Cumulative Distribution Function (CDF) of $log(Chl)$. @fig-model1-priorpc provides

::: {#fig-model1-priorpc fig-pos="H"}

>>>>

This model follows the OC6 formulation proposed by O’Reilly et al. (2019), with the difference that the denominator is the sum of $Rrs555$ and $Rrs670$ rather than the mean. The model mean is a fourth-order polynomial of $log(MBR)$ to estimate $log(Chl)$. The set of proabilistic equations that define this model can be found in the Supplemental Material Section. The model structure and inference flow is represented in @fig-model1-struct as a Directed Acyclic Graph (DAG).

::: {#fig-model1-struct pos=H}
![](images/model1_structure.png)

Model 1 Directed Acyclic Graph showing model parameters and their prior distributions and how they relate . The intercept $α$ and the polynomial slopes, $β1$-$β4$ all have Gaussian priors; $σ$, the likelihood dispersion parameter, receives a Gamma prior; the likelihood itself is constrained by a Truncated Normal distribution (see text for more). The plate containing the likelihood is indexed ($obs\_idx$) indicating the likelihood will be computed for each of the 1082 observations available.

@@

::: {#fig-model2-struct}
![](images/model2_structure.png)
Model 2 - Multi-level hierarchical level with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510). Model parameter prior mean and variance have Gaussian and Exponential hyperpriors. This structure allows the sharing of information across groups at model fitting time.
:::

#### Model 5: HLR with Added Linear Heteroscedastic Likelihood Dispersion
The data plotted in @fig-eda suggests heteroscedasticity - variance that is not constant, but rather changes depending on the data. Models 3 and 4 investigate the role of group-wise constant heteroscedasticity and global linear heteroscedasticity, respectively. Details of these models can be found in the Supplementary Material section. To investigate this further, Model 5 looks at group-wise linear heteroscedasticity. Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood. Similar to the model mean,  $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors. Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.  illustrates the model structure.

::: {#fig-model5-struct}
![](images/model5_structure.png)
Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.  Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2.
:::

####  Model 6: Impact of Chl Measurement Method on Noise
A reasonable assumption is that HPLC measurements are less noisy than fluorescence. This model adds a measurement-method-specific dispersion term to the structure of Model 5, to capture the suspected increase in variability associated with fluorescence-based chlorophyll estimates.  The likelihood dispersion, $σ$, is given the expression $log(σ) = α_j + β_j × log(MBR) + γ_{chl\_type} * (1 - chl_{type\_idx})$, where $chl_{type\_idx} = 1$ if *in-situ* chlorophyll was measured with HPLC, $0$ otherwise. As a result, $σ$ is the same as in Model 5 if $chl_{type\_idx} = 1$, since the extra term is nullified. In the case of fluorescence measurement, on the other hand, $log(σ)$ receives an extra noise term, providing an effective measurement of noise difference between the two methods. @fig-model6-struct shows the inference DAG of Model 6.

::: {#fig-model6-struct}

@@

*Priors* - I used weakly informative priors across all models to balance flexibility with regularization.  For regression parameters  I used Gaussian priors, In the case of the  single-level  Model 1, this meant priors with $0$ mean and variance of $1$ to encourage parameter sparsity. In the case of Models 2-6, which are multi-level models, hyperpriors of prior means were set to Gaussans of mean 0 and variance 1, again to encourage sparsity. Variance hyperpriors were given Exponential distributions with rate set to $1$ with significant density near 0, whichß encourages similarity across groups unless the data mandates otherwise. This hyperprior also encourages uncertainty shrinkage across groups if possible. In the case of homoscedasticity, or group-wise heteroscedasticity, I used a Gamma prior of shape and scale of $2$ for the likelihood dispersion $σ$, which provides strictly postivie support. Prior soundness was assessed using *Prior Predictive Checks*. This procedure takes advantage of the fact taht Bayesian models are generative, meaning the model can be sampled without data, to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting. Prior predictive $log(Chl)$ values were plotted as Cumulative Distribution Function (CDF) of $log(Chl)$. @fig-model1-priorpc shows such a plot for Model 1.

::: {#fig-model1-priorpc}
